# Visualizing data and making reports


```{r echo=FALSE, message=FALSE}
suppressPackageStartupMessages(library(lgrdata))
suppressPackageStartupMessages(library(ggplot2))
suppressPackageStartupMessages(library(ggthemes))
suppressPackageStartupMessages(library(scales))
```

## Introduction

NOT a manual of either, and less detail on both since resources elsewhere are perfect.
But see appendix for base plotting.

We do have a list of DO's and DONT's, one for visualizing data, and one for making reports with rmarkdown.

We also have a list of resources.


### Should you use R for graphics?

- yes because flexible, **reproducible code**
- yes because even if you don't end up using R for publication, report, or fancy presentation - you still need visualization skills when doing analysis
- R graphs were not originally designed for interactivity, but there are many recent inventions (shiny, ggiraph) and ports to Javascript-based graphing tools (plotly, highchartr, d3) 
- R not great at 3D graphs, but as a data analyst you probably should not use many of those (but see rgl!)
- No when you prefer to modify your figure by hand, dragging items, clicking on things to change color etc. Perhaps you have your favorite tool to make figures already.


## Visualizing data {#visualizing}

### Tips

- do a lot of visualizing, all the time, make it a habit rather than what you do at the end. you understand your data this way (and mostly this way)
- most people won't believe your cool results unless it is shown in a figure, **clearly**. I don't care about your tables or your dense statistical results (those you put at the end).
- spend time formatting the figures so they don't look like anyone else's. This gives personal flavor to your visualizations. In ggplot2, you can also just pick a theme you like and stick with it (`ggthemes` link especially), and please do not use the default color scheme or background (also because it is hideous).
- when stuck, quickly google what to do. I do this all the time. Do not read the help files except `?par` for base graphics, otherwise quickly do a web search. Everything can be found (this was a different affair in 2001).


### base or ggplot2? Why not both?

R comes with two graphics systems, that cannot be used together. The `base` graphics system has been part of R from the start, and continues to be used today. The second system is the `grid` system, and the most popular package that uses this system is `ggplot2`, a package in development for over 10 years that now sees very wide application, and numerous add-ons and extensions. 

Personally, I have used base graphics since 2001, and resisted switching to `ggplot2` for a very long time. I now use `ggplot2` for visualizing data, but continue to use base graphics as well, for a few reasons - especially because many add-on packaged include some quick plotting function that returns a base plot.

You can read around the web for many opinionated accounts on why we should use `ggplot2` or the other ([for example here](http://varianceexplained.org/r/why-I-use-ggplot2/)). 

- use ggplot2 if you are a newcomer, easiest to make more complex graphs, reasons include:
  - much easier to make multi-panel, multi-line, multi-everything figures than with base graphics. Very notable examples include multi-panel figures with grouped bar plots, adding regression and smoothed lines *with* confidence bands to plots, overlapping area plots,  histograms and ribbons. 
  - easier to achieve consistent figure formatting with the use of themes
  - lots of graph types are easier to plot than with base (where you often have to find additional packages)
- Many packages include some built-in plotting function, usually in base graphics. It is usually much easier to use those functions directly than to implement the figure from scratch in `ggplot2`. Examples include `car::residualPlot`, `visreg::visreg`, and many others.
- Sometimes base is easier, especially when you are experimenting. To plot a mathematical function, we can do `curve(x^3, from=-1, to=1)` - not so easy in `ggplot2`. Also `ggplot2` has no support for 3D graphs (but we don't use those much anyway), base has quite a lot.
- Use base when you are making a simple plot, and wish to avoid *dependencies*. Keeping dependencies to a minimum will make it more likely your code works in the future, and on someone else's machine.

For the reasons mentioned above, in this book we mostly use ggplot2, except when base graphics are a lot easier (that is, use less code).


### base graphics

In base graphics, we can choose from the following main plotting functions (this list is not complete, especially if you consider add-on packages):

+------------------+------------------------------------------------+
| Function         | Key functions                                   |
+==================+================================================+
| `plot`           |       Scatter plots, line plots, various others.        |
+------------------+------------------------------------------------+
| `barplot`        |    Bar plot (including stacked and grouped bar plots)                  |
+------------------+------------------------------------------------+
| `hist`           |      Histograms and (relative) frequency diagrams                |
+------------------+------------------------------------------------+
| `curve`          |       Curves of mathematical expressions               |
+------------------+------------------------------------------------+
| `pie`            |       Pie charts (for less scientific uses)               |
+------------------+------------------------------------------------+
| `boxplot`        |    Box-and-whisker plots                  |
+------------------+------------------------------------------------+
| `symbols`        |   Like scatter plot, but symbols are sized by another variable                   |
+------------------+------------------------------------------------+

Here it is useful to know that `plot` is a so-called *generic function*, which has *methods* for many special objects in R. For example, if we fit a linear regression:

```{r eval=FALSE}
lmfit <- lm(disp ~ hp, data = mtcars)
```

We can use the very brief command `plot(lmfit)` to produce four diagnostic plots. This works because a function called `plot.lm` is actually used when you send an `lm` object to the `plot` function (thus we can look at useful options in this case in the help page `?plot.lm`).

Similarly, `plot` will make different plots based on what type of data you use. In Fig. \@ref(fig:plotdefault) four examples are shown; each of these is the result from a call to `plot()`, but with different data types.

```{r plotdefault, fig.width=3.5, fig.height=3.5, echo=FALSE, fig.cap="Four possible outcomes of a basic call to plot(), depending on whether the Y and X variables are numeric or factor variables. The term 'numeric dataframe' means a dataframe where all columns are numeric.", fig.show='hold'}
data(allometry)
data(titanic)

par(cex=0.9, cex.lab=0.9, cex.axis=0.8, cex.main=0.9, font.main=3)
plot(height ~ diameter, data=allometry, main="numeric vs. numeric")
plot(diameter ~ species, data=allometry, main="numeric vs. factor")
plot(allometry[,2:4], main="numeric dataframe")
plot(as.factor(Survived) ~ PClass, data=titanic, ylab="Survived", main="factor vs. factor")
```



#### Resources

- appendix



### ggplot2 {#ggplot2}

The `ggplot2` package is an incredibly rich and powerful toolkit to construct plots. It is far beyond the scope of this book to provide a complete overview of the package, but luckily lots of free online resources are available (see Section \@ref(ggplotresources)).

We will show the functionality of `ggplot2` by providing two examples, one scatterplot, and one barplot. Since we use `ggplot2` throughout this book, you can find many more examples. The following three examples were chosen because they highlight a few key options.


#### Formatting data for `ggplot2`

To use `ggplot2`, our data have to always be in *long format*, where each row represents an observation, and each column in the dataframe represents a type of measurement. Most of the datasets included in the `lgrdata` package are already in long format. One exception is the `dutchelection` dataset:

```{r}
data(dutchelection)
head(dutchelection, 3)
```

We cannot really use `ggplot2` when the data are formatted like this. Instead we want to have a dataframe with columns 'Date', 'party', and 'vote_percent', so that we can label the graph by political party. As we will see in Section \@ref(), quite magically we can use `melt` from `reshape2` :

```{r message=FALSE}
library(reshape2)
dutchlong <- melt(dutchelection, variable.name = "party", value.name="vote_percent")
head(dutchlong, 3)
```



#### Scatterplot example




#### Area plot example


```{r}
data(oil)
ggplot(oil, aes(x=year, y=production/1000, fill=country)) + 
  geom_area() + 
  labs(x = "", y = "Crude Oil Production (MOE)") +
  scale_fill_economist() + 
  theme_tufte()
```



#### Barplot example

More complex barplots is one area where `ggplot2` is much, much easier to use than base graphics. If you want to represent count data as bars - a common application of barplots - `ggplot2` does the counting for you automatically, saving a step where such tables of counts have to be constructed (as in Section \@ref()). 

If instead you want to make a barplot of data *as they are* (that is, don't make tables first), use the setting `geom_bar(stat = "identity")`, instead of `stat = "count"`, as in the example below.

In the following example we use the `titanic` data, available in the `lgrdata` package. 

```{r}
library(lgrdata)
data(titanic)

# Important to convert factor variables to factors!
# 'Survived' is coded as 0,1 - but this is really a factor.
titanic$Survived <- as.factor(titanic$Survived)

library(ggplot2)
library(ggthemes)
```

The following code is the minimal code needed to make a barplot with two panels (one for female passengers, one for male), counting passengers that have survived or died, for each of the three passenger classes.

```{r eval = FALSE}
# Result not shown - see next example
ggplot(titanic, aes(x=PClass, fill=Survived)) +
  geom_bar(stat="count", position=position_dodge()) +
  facet_wrap(~Sex)
```

You can run the code above yourself to see what is produced. It does not look great, so instead we re-format the legend, the colors, etc., and end up with Fig. \@ref(fig:titanicplot1). For both of these figures, *make sure that `Survived` is a factor variable*, otherwise the plot won't look right.

```{r titanicplot1}
ggplot(titanic, aes(x=PClass, fill=Survived)) +        # bars by passenger class,
                                                       # split by 'Survived'  
  scale_fill_manual(labels=c("Died", "Survived"),      # labels for legend
                    values=c("red2","grey")) +         # colors for bars (and legend)
  geom_bar(stat="count", position=position_dodge()) +  # count data, bars side by side
  facet_wrap(~Sex) +                                   # panels by Sex
  theme_tufte() +                                      # a very minimal theme
  theme(legend.position = "right",                     # legend position (default)
        legend.title=element_blank())                  # no legend title
```


Same figure but stacked bars, expressed as percent of total (Fig. \@ref(fig:titanicplot2)).


```{r titanicplot2}
ggplot(titanic, aes(x=PClass, fill=as.factor(Survived))) +
  scale_fill_manual(labels=c("Died", "Survived"),      # labels for legend
                    values=c("red2","grey")) +         # colors for bars (and legend)
  geom_bar(stat="count", position=position_fill()) +   # count the data, stack the bars
  scale_y_continuous(labels=scales::percent) +         # express y-axis as % 
  facet_wrap(~Sex) +                                   # two panels for each sex
  theme_tufte() +                                      # a very minimal theme
  theme(legend.position = "right",                     # legend position (default)
        legend.title=element_blank())                  # no legend title
```


The `ggplot2` package has many extensions, packages that offer additional options, themes, plotting layouts, etc. In this book, we only use a couple of add-on packages for reasons of brevity. These include:

- `ggthemes` : an excellent collection of themes and color palettes
- `scales` : useful functions for formatting axes, adjusting colors
- `gridExtra` : to arrange multiple, separate plots, on one page 


#### Resources {#ggplotresources}

- my favorite: http://www.cookbook-r.com/Graphs/, lots of small examples
- also very practical gallery guide at http://www.sthda.com/english/wiki/ggplot2-essentials
- official page at https://ggplot2.tidyverse.org/, see cheat sheet
- more gentle introduction at https://r4ds.had.co.nz/data-visualisation.html
- nice gallery of useful plots, with all code at http://r-statistics.co/Top50-Ggplot2-Visualizations-MasterList-R-Code.html




## Making reports

### rmarkdown {#rmarkdown}

Script files are good for simple analyses, and they are great for storing small amounts of code that you would like to use in lots of different projects. They are also the right solution if you want to produce R scripts that are executed as part of a larger project, for example from the command line, or executed on a schedule (e.g. daily data quality checking). In Section \@ref(scripts) we give a few more tips.

But, scripts are not the best way to share your results with others. Instead, we strongly suggest you use R markdown - a document format that includes both text with markup, and *code chunks*, bits of R code that is executed, and the results placed in the document. 

Markdown is a simple set of rules for formatting text files so that they are both human-readable and processable by software, and is used not just by `rmarkdown` but in many applications (like Github and Stackoverflow). R Markdown simply adds R, by allowing code chunks with various options.

RStudio offers a handy editor for markdown files, and starts with an example file that shows many of the features. Start a new markdown file by choosing `File > New File > R Markdown...`. For now, leave the default settings. The new R markdown document (which is just a text file with the extension `.Rmd`) already contains some example code. Run this example by clicking the button just above the markdown document (Knit HTML). The example code will generate a new document (an HTML page), which opens in your browser You can see that the output contains text, R code, and even a plot.

Throughout this book we assume that you use `rmarkdown` for all your reporting needs. It is by far the best option available at the moment, and provides many options for customization.



### Resources

The documentation for `rmarkdown` based documents is a little scattered across various sources. The reason for this is that multiple packages are involved in making an HTML page from an `.Rmd` file. 

- The `knitr` package handles the *execution* of the R code in the document, and the *integration* of the R output into the document. It provides all options that are set at the top of each code chunk, for example to show or hide the code, to control figure size and resolution, captions for tables. All chunk options are [listed on the `knitr` website](https://yihui.name/knitr/options/).
- The `rmarkdown` package (from Rstudio) handles the *conversion* to the desired output format, for example HTML or PDF. To do so, it uses Pandoc (see below), which is installed automatically with Rstudio. The `rmarkdown` package provides all options at the top of the `Rmd` document (the so-called YAML header), including the desired output format, and to control many options like the table of contents, format style, custom CSS, and so on. All options for `rmarkdown` [are described on Rstudio's website](https://rmarkdown.rstudio.com/). (Make sure to scroll down and select an output format to read about available options).
- Finally, Pandoc is the actual engine to convert our raw text file (including R output formatted by `knitr`) to an HTML page, or some other format. Normally you don't have to learn much about Pandoc's options, since many settings are already provided by `rmarkdown`, but [the various formatting rules and conversion options are described here](https://pandoc.org/MANUAL.html).

