# Appendix: catalogue of self-starting functions for non-linear regression {#nonlinearappendix}


```{r, echo=FALSE}
ss_plot_i <- function(model, params, xlim, add=FALSE, inputname="input", ...){
  
  x <- seq(xlim[1], xlim[2], length=501)
  li <- list(input=x)
  names(li)[1] <- inputname
  l <- c(li, params)
  y <- do.call(model, l)
  
  if(!add){
    plot(x,y, type='l', ...)
  } else {
    lines(x,y, ...)
  }
}
make_legend <- function(p){
  n <- length(p[[1]])
  txt <- c()
  for(i in 1:n){
    txt[i] <- paste(paste(names(p), "=", sapply(p, "[", i)), collapse=",")
  }
  txt
}
ss_plot <- function(model, params, legend="bottomright", ...){
  n <- length(params[[1]])
  
  par(xpd=FALSE)
  fn <- function(i, ...)ss_plot_i(model, lapply(params, "[", i), ...)
  
  fn(1, col=palette()[1], ...)
  for(j in 2:n){
    fn(j, add=TRUE, col=palette()[j],...)
  }
  
  txt <- make_legend(params)
  par(xpd=TRUE)
  legend("bottom", txt, lty=1, col=palette(), bty='n', cex=1.1, inset=-0.25)
}
get_help_title <- function(ss){
  rdbfile <- file.path(find.package("stats"), "help", "stats")
  rdb <- tools:::fetchRdDB(rdbfile, key = ss)
  rdb[[1]][[1]][1]
}

palette(c("blue","red","forestgreen"))
```

## SSasymp

`r get_help_title("SSasymp")`


$$Y = Asym+(R0-Asym)*exp(-exp(lrc)*input)$$

```{r, fig.width=9, fig.height=5, echo=FALSE}
par(mfrow=c(1,3), mar=c(8,4,0.5,1), mgp=c(2,0.5,0), tcl=0.2, cex.axis=0.8, cex.lab=1.1,
    oma=c(0,0,2,0))
ss_plot(SSasymp, list(Asym=c(10,10,10), R0=c(6,4,2), lrc=c(0.1,0.1,0.1)), xlim=c(0,6), ylim=c(0,10))
ss_plot(SSasymp, list(Asym=c(10,10,10), R0=c(2,2,2), lrc=c(3,1,0.025)), xlim=c(0,6), ylim=c(0,10))
ss_plot(SSasymp, list(Asym=c(2,7,10), R0=c(4,4,4), lrc=c(1,1,1)), xlim=c(0,6), ylim=c(0,10))
```

## SSasympOff
`r get_help_title("SSasympOff")`

$$Y = Asym*(1 - exp(-exp(lrc)*(input - c0)))$$

```{r, fig.width=9, fig.height=5, echo=FALSE}
par(mfrow=c(1,3), mar=c(8,4,0.5,1), mgp=c(2,0.5,0), tcl=0.2, cex.axis=0.8, cex.lab=1.1,
    oma=c(0,0,2,0))
ss_plot(SSasympOff, list(Asym=c(10,10,10), c0=c(1,2,3), lrc=c(0.1,0.1,0.1)), "topleft", xlim=c(0,10), ylim=c(0,12))
ss_plot(SSasympOff, list(Asym=c(6,8,10), c0=c(1,1,1), lrc=c(1,1,1)), xlim=c(0,6), ylim=c(0,10))
ss_plot(SSasympOff, list(Asym=c(10,10,10), c0=c(1,1,1), lrc=c(0.1,1,2)), xlim=c(0,6), ylim=c(0,10))
```


## SSasympOrig
`r get_help_title("SSasympOrig")`

$$Y = Asym*(1 - exp(-exp(lrc)*input))$$

```{r, fig.width=9, fig.height=5, echo=FALSE}
par(mfrow=c(1,3), mar=c(8,4,0.5,1), mgp=c(2,0.5,0), tcl=0.2, cex.axis=0.8, cex.lab=1.1,
    oma=c(0,0,2,0))
ss_plot(SSasympOrig, list(Asym=c(6,8,10), lrc=c(1,1,1)), xlim=c(0,6), ylim=c(0,10))
ss_plot(SSasympOrig, list(Asym=c(10,10,10), lrc=c(0.1,1,2)), xlim=c(0,6), ylim=c(0,10))
```


## SSbiexp
`r get_help_title("SSbiexp")`

$$Y = A1*exp(-exp(lrc1)*input)+A2*exp(-exp(lrc2)*input)$$

Use this model when a simple exponential model ($A1*exp(-exp(lrc1)*input)$) doesn't quite fit. For the simple exponential model, you won't need a self-starting function as even poor starting values will result in convergence. Also, it can be linearized after taking the log of both the predictor and the response variable. 

```{r, fig.width=9, fig.height=5, echo=FALSE}
par(mfrow=c(1,3), mar=c(8,4,0.5,1), mgp=c(2,0.5,0), tcl=0.2, cex.axis=0.8, cex.lab=1.1,
    oma=c(0,0,2,0))
ss_plot(SSbiexp, list(A1=c(5,5,5), lrc1=c(1,1,1), A2=c(1,5,10), lrc2=c(0.1,0.1,0.1)), "topright", 
        xlim=c(0,10), ylim=c(0,5))
ss_plot(SSbiexp, list(A1=c(1,1,1), lrc1=c(0.005,0.2,0.5), A2=c(-1,-1,-1), lrc2=c(2,2,2)), "topright", 
        xlim=c(0,5), ylim=c(-1,1))

```


## SSmicmen

`r get_help_title("SSmicmen")`

$$ Y = Vm*input/(K+input)$$


```{r, fig.width=9, fig.height=5, echo=FALSE}
par(mfrow=c(1,3), mar=c(8,4,0.5,1), mgp=c(2,0.5,0), tcl=0.2, cex.axis=0.8, cex.lab=1.1,
    oma=c(0,0,2,0))
ss_plot(SSmicmen, list(Vm=c(200, 200, 200), K=c(0.05, 0.1, 0.2)), xlim=c(0,1), ylim=c(0,200))
ss_plot(SSmicmen, list(Vm=c(100, 150, 200), K=c(0.1, 0.1, 0.1)), xlim=c(0,1), ylim=c(0,200))
```


## SSlogis

`r get_help_title("SSlogis")`

$$Y = Asym/(1+exp((xmid-input)/scal))$$


```{r, fig.width=9, fig.height=5, echo=FALSE}
par(mfrow=c(1,3), mar=c(8,4,0.5,1), mgp=c(2,0.5,0), tcl=0.2, cex.axis=0.8, cex.lab=1.1,
    oma=c(0,0,2,0))
ss_plot(SSlogis, list(Asym=c(100,100,100), xmid=c(10,20,30), scal=c(2,2,2)), xlim=c(0,60), ylim=c(0,100))
ss_plot(SSlogis, list(Asym=c(50,75,100), xmid=c(20,20,20), scal=c(2,2,2)), xlim=c(0,60), ylim=c(0,100))
ss_plot(SSlogis, list(Asym=c(100,100,100), xmid=c(20,20,20), scal=c(1,2,4)), xlim=c(0,60), ylim=c(0,100))
```


## SSgompertz

`r get_help_title("SSgompertz")`

$$Y = Asym*exp(-b2*b3^x)$$

Note: the parameter $b3$ should be less than 1.

```{r, fig.width=9, fig.height=5, echo=FALSE}
par(mfrow=c(1,3), mar=c(8,4,0.5,1), mgp=c(2,0.5,0), tcl=0.2, cex.axis=0.8, cex.lab=1.1,
    oma=c(0,0,2,0))
ss_plot(SSgompertz, list(Asym=c(40,50,60), b2=c(2,2,2), b3=c(0.8, 0.8, 0.8)),
        xlim=c(0,30), ylim=c(0,60), inputname="x")
ss_plot(SSgompertz, list(Asym=c(50,50,50), b2=c(1,2.5,5), b3=c(0.8, 0.8, 0.8)),
        xlim=c(0,30), ylim=c(0,60), inputname="x")
ss_plot(SSgompertz, list(Asym=c(50,50,50), b2=c(5,5,5), b3=c(0.4, 0.7, 0.8)),
        xlim=c(0,30), ylim=c(0,60), inputname="x")
```

## SSweibull

`r get_help_title("SSweibull")`

$$Y = Asym-Drop*exp(-exp(lrc)*x^pwr)$$

Note: in the plot below, the \code{Asym} parameter is not varied, but as before, it represents the asymptote.

```{r, fig.width=9, fig.height=5, echo=FALSE}
par(mfrow=c(1,3), mar=c(8,4,0.5,1), mgp=c(2,0.5,0), tcl=0.2, cex.axis=0.8, cex.lab=1.1,
    oma=c(0,0,2,0))
ss_plot(SSweibull, list(Asym=c(150,150,150), Drop=c(50,100,150), lrc=c(-6, -6, -6), pwr=c(2.5, 2.5, 2.5)),
        inputname="x",xlim=c(0,30), ylim=c(0,150))
ss_plot(SSweibull, list(Asym=c(150,150,150), Drop=c(150,150,150), lrc=c(-2, -4, -6), pwr=c(2.5, 2.5, 2.5)),
        inputname="x",xlim=c(0,30), ylim=c(0,150))
ss_plot(SSweibull, list(Asym=c(150,150,150), Drop=c(150,150,150), lrc=c(-6, -6, -6), pwr=c(2.2, 2.5, 2.8)),
        inputname="x",xlim=c(0,30), ylim=c(0,150))

```


## SSfpl


`r get_help_title("SSfpl")`

$$Y = A+(B-A)/(1+exp((xmid-input)/scal))$$


```{r, fig.width=9, fig.height=5, echo=FALSE}
par(mfrow=c(1,3), mar=c(8,4,0.5,1), mgp=c(2,0.5,0), tcl=0.2, cex.axis=0.8, cex.lab=1.1,
    oma=c(0,0,2,0))
ss_plot(SSfpl, list(A=c(0,5,10), B=c(20,20,20), xmid=c(20,20,20), scal=c(2,2,2)), 
        xlim=c(0,50), ylim=c(0,20))
ss_plot(SSfpl, list(A=c(5,5,5), B=c(10,15,20), xmid=c(20,20,20), scal=c(2,2,2)), 
        xlim=c(0,50), ylim=c(0,20))
ss_plot(SSfpl, list(A=c(5,5,5), B=c(20,20,20), xmid=c(10,20,30), scal=c(2,2,2)), 
        xlim=c(0,50), ylim=c(0,20))
```


```{r, fig.width=9, fig.height=5, echo=FALSE}
par(mfrow=c(1,3), mar=c(8,4,0.5,1), mgp=c(2,0.5,0), tcl=0.2, cex.axis=0.8, cex.lab=1.1,
    oma=c(0,0,2,0))
ss_plot(SSfpl, list(A=c(5,5,5), B=c(20,20,20), xmid=c(20,20,20), scal=c(1,2,4)), 
        xlim=c(0,50), ylim=c(0,20))

```


