
### Working with a single vector - indexing

Recall Exercise \@ref(vecexerc1). Load the `rainfall` data once more.

```{r echo=FALSE}
rainfall <- c(0.1,0.6, 33.8,  1.9,  9.6,  4.3, 33.7,  0.3,  0.0,  0.1)
```


We now practice subsetting a vector (see Section \@ref(vectorindexing)).

**1.** Take a subset of the rainfall data where rain is larger than 20.
```{r }
rainfall[rainfall > 20]
```

 **2.** What is the mean rainfall for days where the rainfall was at least 4?
```{r }
mean(rainfall[rainfall >= 4])
```

 **3.** Subset the vector where it is either exactly zero, or exactly 0.6. 
```{r }
rainfall[rainfall == 0 | rainfall == 0.6]

# Alternative solution,
rainfall[rainfall %in% c(0, 0.6)]
```



### Alphabet aerobics 2

The 26 letters of the Roman alphabet are conveniently accessible in R via `letters` and `LETTERS`. These are not functions, but vectors that are always loaded.


**1.** What is the 18th letter of the alphabet?

```{r }
# You could have guessed,
LETTERS[18]

# or letters[18]
```

**2.** What is the last letter of the alphabet (pretend you don't know the alphabet has 26 letters)?

```{r }
# Use length() to index the vector,
letters[length(letters)]

# You could use you knowledge that the alphabet contains 26 letters, but the above 
# solution is more general.
```


**3.** Use `?sample` to figure out how to sample with replacement. Generate a random subset of fifteen letters. Are any letters in the subset duplicated? *Hint:* use the `any` and `duplicated` functions. Which letters?

```{r }
# 1. Random subset of 15 letters
let15 <- sample(letters, 15, replace = TRUE)

# 2. Any duplicated?
any(duplicated(let15))

# 3. Which are duplicated?
# This tells us the index of the letter that is replicated,
which(duplicated(let15))  

# We can use it to index letters to find the actual letters
let15[which(duplicated(let15))]
```



### Basic operations with the Cereals data

First read or otherwise understand Section \@ref(vecdataframes).

For this exercise, we will use the `cereals` dataset (from the `lgrdata` package, use `data(cereals)` to load it).

**1.** Read in the dataset, look at the first few rows with `head` and inspect the data types of the variables in the dataframe with `str`.
```{r }
library(lgrdata)
data(cereals)

str(cereals)
head(cereals)
```


**2.** Add a new variable to the dataset called `totalcarb`, which is the sum of `carbo` and `sugars`. You can do this is in (at least) three ways! (One of which is `mutate` from `dplyr` - a good idea to start using this powerful function).

```{r }
# Option 1.
cereals$totalcarb <- cereals$carbo + cereals$sugars

# Alternatively:
cereals$totalcarb <- with(cereals, carbo + sugars)

# Alternatively:
cereals <- mutate(cereals, 
                  totalcarb = carbo + sugars)

```

**3.** How many cereals in the dataframe are 'hot' cereals? Either make a subset and count the rows, or subset the `Cold.or.Hot` vector directly, and determine its `length` (do not use `table` yet, or a similar function!).
 
```{r }
# Solution 1 : using subset on the dataframe
hotcereals <- subset(cereals, Cold.or.Hot == "H")

# Or (faster for large datasets):
library(dplyr)
hotcereals <- filter(cereals, Cold.or.Hot == "H")

# how many rows?
nrow(hotcereals)

# Solution 2 : indexing the vector 'Cold.or.Hot'
hot <- cereals$Cold.or.Hot[cereals$Cold.or.Hot == "H"]
length(hot)

```

**4.** How many unique manufacturers are included in the dataset? *Hint:* use `length` and `unique`.

```{r }
length(unique(cereals$Manufacturer))
```

**5.** Now try the `n_distinct` function from `dplyr`.


**6.** Take a subset of the dataframe of all cereals that have less than 80 calories,
AND have more than 20 units of vitamins.
```{r }
subset(cereals, calories < 80 & vitamins > 20)

# OR
# Note how with dplyr we can use a comma to separate AND (&)
# statements. It is also much faster for larger datasets.
dplyr::filter(cereals, 
              calories < 80, 
              vitamins > 20)
```


**7.** Take a subset of the dataframe containing cereals that contain at least 1 unit of sugar, and keep only the variables 'Cereal.name', 'calories' and 'vitamins'. Then inspect the first few rows of the dataframe with `head`.


```{r }
# Using 'subset' (base R)
cereal_subs <- subset(cereals, 
                      sugars >= 1, 
                      select=c(Cereal.name, calories, vitamins))

# dplyr:
library(dplyr)
cereal_subs <- filter(cereals, 
                      sugars >= 1) %>%
               dplyr::select(Cereal.name, calories, vitamins)


# You can also use square bracket indexing.
# In practice you will rarely need this, but it is good to understand
# how it works:
cereal_subs <- cereals[cereals$sugars >= 1, c("Cereal.name","calories","vitamins")]

# In the above, '>=' means 'larger than or equal to'.

# Look at first few rows of the new subset.
head(cereal_subs)
```


**8.** For one of the above subsets, write a new CSV file to disk using `write.csv` (see Section \@ref(exportingdata)).

```{r eval=FALSE}
# Use row.names=FALSE to avoid unnecessary row numbers in the CSV file.
write.csv(cereal_subs, "cerealsubset.csv", row.names=FALSE)
```


**9.** Rename the column 'Manufacturer' to 'Producer' (see Section  \@ref(namesdataframe)).
```{r eval=FALSE}
# First look at the names:
names(cereals)

# So, the second name is Manufacturer. Change it:
names(cereals)[2] <- "Producer"

# for dplyr, we need to know what the current name is:
cereals <- rename(cereals, 
                  Producer = Manufacturer)

```






### A short dataset


**1.** Read the following data into R (number of cuckoos seen in a week by an avid birdwatcher). Give the resulting dataframe a reasonable name. *Hint:*To read this dataset, look at Section  \@ref(datainscript) for a possibility (there are at least two ways to read this dataset, or you can type it into Excel and save as a CSV file if you prefer).

```
Day nrbirds
sunday 3
monday 2
tuesday 5
wednesday 0
thursday 8
friday 1
saturday 2
```


```{r readhoney, tidy=FALSE}
cuckoos <- read.csv(text="
Day, nrbirds
sunday, 3
monday, 2
tuesday, 5
wednesday, 0
thursday, 8
friday, 1
saturday, 2")
```


**2.** Add a day number to the dataset you read in above (sunday=1, saturday=7). Recall the `seq` function (Section  \@ref(sequences)).
```{r honeyadd}
cuckoos$daynumber <- 1:7
```

**3.** Delete the 'Day' variable (to only keep the `daynumber` that you added above). 
```{r }
# Solution 1
cuckoos2 <- subset(cuckoos, select=-Day)

# Solution 2 (simply deletes the first column)
cuckoos2 <- cuckoos[,-1]

# Solution 3
cuckoos2 <- dplyr::select(cuckoos, -Day)
```

```{r echo=FALSE}
<<readhoney>>
<<honeyadd>>
```

**4.** On which `daynumber` did you observe the most honeyeaters? *Hint:* use `which.max`, in combination with indexing.

```{r }
cuckoos$daynumber[which.max(cuckoos$nrbirds)]
```



**5.** Sort the dataset by number of birds seen. *Hint:* use the `order` function to find the order of the number of birds, then use this vector to index the dataframe.
```{r }
nrbird_order <- order(cuckoos$nrbirds)

honey_sorted <- cuckoos[nrbird_order,]
```




### Titanic - Part 1


**1.** Read the `titanic` data from the `lgrdata` package.
 
```{r }
data(titanic)
```


**2.** Make two new dataframes : a subset of male survivors, and a subset of female survivors. Recall Section  \@ref(subsetdataframes).  Use `filter` from `dplyr`.
```{r }
# Always first inspect your data to see what it looks like
head(titanic)

# Probably already loaded!
library(dplyr)

# Then take subsets
titanic_male <- filter(titanic, 
                       Sex == "male", 
                       Survived == 1)
titanic_female <- filter(titanic, 
                       Sex == "female", 
                       Survived == 1)

```


**3.** Based on the previous question, what was the name of the oldest surviving male? In what class was the youngest surviving female? *Hint:* use `which.max`, `which.min` on the subsets you just created. 

The easiest solution here is to use square bracket notation, but you can also solve this question in steps.

```{r }
# Base R has the shortest solution:
titanic_male$Name[which.max(titanic_male$Age)]

titanic_female$PClass[which.min(titanic_female$Age)]

```


**4.** Take 15 random names of passengers from the Titanic, and sort them alphabetically. *Hint:* use `sort`.
```{r }
sample15 <- sample(titanic$Name, 15)
sort(sample15)
```



### Titanic - Part 2
 
**1.** Convert the 'Name' (passenger name) variable to a 'character' variable, and store it in the dataframe. See Section  \@ref(textgrep).
```{r }
titanic$Name <- as.character(titanic$Name)
```

**2.** How many observations of 'Age' are missing from the dataframe? See examples in Section  \@ref(workingmissing).
```{r }
# Look at summary:
summary(titanic$Age)

# Or count the number of missing values directly.
sum(is.na(titanic$Age))
```

**3.** Make a new variable called 'Status', based on the 'Survived' variable already in the dataset. For passengers that did not survive, Status should be 'dead', for those who did, Status should be 'alive'. Make sure this new variable is a factor. See the example with the `ifelse` function in Section  \@ref(workingfactors).
```{r }
# ifelse may return a factor, but to be sure we also use factor().
titanic$Status <- factor(ifelse(titanic$Survived==0,"dead","alive"))
```

**4.** Count the number of passengers in each class (1st, 2nd, 3rd). *Hint:* use `table` as shown in Section  \@ref(workingfactors).
```{r }
# The easiest way to count number of observations for each level of a factor is table()
table(titanic$PClass)
```

**5.** Using `grep`, find the six passengers with the last name 'Fortune'. Make this subset into a new dataframe. Did they all survive? *Hint:* to do this, make sure you recall how to use one vector to index a dataframe (see Section  \@ref(subsetdataframes)). Also, the `all` function might be useful here (see Section  \@ref(workinglogic)).
```{r }
# Solution with 'grepl' (which returns a vector of TRUE/FALSE)
fortunepass <- subset(titanic, grepl("Fortune", Name))

# Did they *all* survive?
all(fortunepass$Survived == 1)
```

**6.** As in *2.*, for what proportion of the passengers is the age unknown? Was this proportion higher for 3rd class than 1st and 2nd? *Hint:* First make a subset of the dataframe where age is missing (see Section \@ref(naindataframe)), and then use `table`, as well as `nrow`.
```{r tidy=TRUE}
# Solution 1
# First subset the data to extract only the ones with missing Age.
# Then, count the number of observations by PClass
titanic_missage <- subset(titanic, is.na(Age))
table(titanic_missage$PClass)

# Finally, divide by the total number of rows to get
# the proportions:
table(titanic_missage$PClass) / nrow(titanic_missage)
```




### Hydro dam

Use the `hydro` dam data (used in Section \@ref(datesdataframe)).

**1.** Start by reading in the data. Change the first variable to a `Date` 
class (see Section \@ref(readingdates)). 

```{r }
data(hydro)

# Note that the format of Date was D/M/Y.
library(lubridate)
hydro$Date <- dmy(hydro$Date)
```


**2.** Are the successive measurements in the dataset always exactly one week apart? *Hint:* use `diff`.
```{r }
# diff() gives us the sequential differences, we can then list the unique values
# of these differences.
unique(diff(hydro$Date))
```

**3.** Assume that a dangerously low level of the dam is 235 $Gwh$. How many weeks was the dam level equal to or lower than this value?
```{r }
# Because the answer to the previous question was yes, 
# we can just count the number of observations where storage was < 235, like so:
sum(hydro$storage < 235)
```


**4.** (**Hard question**). For question *2.*, how many times did `storage` decrease below 235 (regardless of how long it remained below 235)? *Hint:* use `diff` and `subset`).
```{r }
# SOLUTION 1
# Take a subset of the data where storage < 235
hydrolow <- subset(hydro, storage < 235)

# Look at time difference between successive dates
diff(hydrolow$Date)

# whenever this time difference is larger than 7,
# the level has dipped below 235 again
# (plus one, because the hydrolow dataset starts below 235)
sum(diff(hydrolow$Date) > 7) + 1

# SOLUTION 2
# Add variable that is 0 when storage < 235, 
# 1 otherwise:
hydro$storageBinary <- ifelse(hydro$storage< 235,0,1)

# Now, diff gives -1 when storage dipped below 235:
diff(hydro$storageBinary)

# How many times did it dip below 235?
difs <- diff(hydro$storageBinary)
sum(difs == -1)
```




### HFE tree measurements

Use the data for an experiment where trees were irrigated and/or fertilized (the `hfeplotmeans` dataset).

**1.** Read the data, write a copy to a new object called `trees` (easier to type!) and look at various summaries of the dataset. Use `summary`, `str` and `describe` (the latter is in the `Hmisc` package).
```{r eval=FALSE}
data(hfeifplotmeans)
trees <- hfeifplotmeans

summary(trees)
str(trees)

# Instead of library(Hmisc), which avoids some conflicts
Hmisc::describe(trees)

```
```{r echo=FALSE}
data(hfeifplotmeans)
trees <- hfeifplotmeans
```


**2.** From these summaries, find out how many missing values there are for `height` and `diameter`. Also count the number of missing values as shown in Section  \@ref(workingmissing).
```{r }
# is.na() gives a vector of TRUE/FALSE, which we can sum because TRUE is coded as 1,
# FALSE coded as 0.
sum(is.na(trees$height))
sum(is.na(trees$diameter))
```

**3.** Inspect the levels of the treatment (`treat`), with the `levels` function. Also count the number of levels with the `nlevels` function. Now assign new levels to the factor, replacing the abbreviations with a more informative label. Follow the example in Section  \@ref(changelevels).

```{r }
# Inspect levels
levels(trees$treat)

# Count levels
nlevels(trees$treat)

# Replace levels
levels(trees$treat) <- c("Control","Fertilized","Irrigated","Liquid fertilizer")
```

**4.** Using `table`, count the number of observations by `treat`, to check if the dataset is balanced. Be aware that `table` simply counts the number of rows, regardless of missing values. Now take a subset of the dataset where `height` is not missing, and check the number of observations again.

```{r }
# Count by factor levels
table(trees$treat)

# Take subset of non-missing data and try again
trees_nona <- subset(trees, !is.na(height))
table(trees_nona$treat)
```


**5.** For which dates do missing values occur in `height` in this dataset? *Hint:* use a combination of `is.na` and `unique`.
```{r }
# First make it a Date class:
library(lubridate)
trees$Date <- mdy(trees$Date)

# Then find unique Dates where height was NA:
unique(trees$Date[is.na(trees$height)])

# Or, alternatively:
unique(trees[is.na(trees$height), "Date"])
```





### Flux data

In this exercise, you will practice working with Dates and Date-Time combinations, with a timeseries dataset (the `fluxtower` dataset from the `lgrdata` package). 

In this dataset, a new row was produced every 30min - of various meteorological measurements above a forest in Spain. For example, `FCO2` is the flux of carbon dioxide out of forest, so that negative values indicate photosynthesis.

**1.** Read the dataframe. Rename the first column to 'DateTime' (recall Section  \@ref(namesdataframe)).
```{r }
data(fluxtower)

# Rename the first column to 'DateTime'.
names(fluxtower)[1] <- "DateTime"
```


**2.** Convert DateTime to a `POSIXct` class. Beware of the formatting (recall Section  \@ref(datetime)).

```{r }
library(lubridate)

# Note the format was in D/M/Y H:M.
fluxtower$DateTime <- dmy_hm(fluxtower$DateTime)
```


**3.** Did the above action produce any missing values? Were these already missing in the original dataset?

```{r }
# is.na returns a vector of TRUE/FALSE, any() will be TRUE if at least one of the
# supplied values is TRUE.
any(is.na(fluxtower$DateTime))
```

**4.** Add a variable to the dataset called 'Quality'. This variable should be 'bad' when the variable 'ustar' is less than 0.15, and 'good' otherwise. Recall the example in Section \@ref(workingfactors).
```{r }
fluxtower$Quality <- as.factor(ifelse(fluxtower$ustar < 0.15, "bad","good"))
```


**5.** Add a 'month' column to the dataset, as well as 'year'. 
```{r }
# Abbreviated month : see ?month for more options
fluxtower$month <- month(fluxtower$DateTime, label=TRUE)

# Year
fluxtower$year <- year(fluxtower$Date)
```


**6.** Look at the 'Rain' column. There are some problems; re-read the data or find another way to display `NA` whenever the data have an invalid value. *Hint:* look at the argument `na.strings` in `read.table`. 

Did it rain on this forest in Spain?
```{r }
# str() shows us that rain is not a numeric variable as one might expect:
str(fluxtower$Rain)

# Solution 2: set bad values to NA
fluxtower$Rain[fluxtower$Rain == "#DIV/0!"] <- NA
```




<!-- ### Alphabet Aerobics 3 -->

<!-- In this exercise you will practice a bit more working with text, using the lyrics of the song 'Alphabet Aerobics' by Blackalicious. The lyrics are provided as a text file, which we can most conveniently read into a vector with `readLines`, like this, -->
<!-- \begin{verbatim} -->
<!-- lyric <- readLines("alphabet.txt") -->
<!-- \end{verbatim} -->


<!--  **1.** Read the text file into a character vector like above. Count the number of characters in each line (*Hint* : use `nchar`). -->
<!-- ```{r } -->
<!-- lyric <- readLines("alphabet.txt") -->
<!-- nchar(lyric) -->
<!-- ``` -->

<!-- **1.** Extract the first character of each line (recall examples in Section  \@ref(workingwithtext} on p.~ \@ref(workingwithtext}), and store it in a vector. Now sort this vector alphabetically and compare it to the unsorted vector. Are they the same? (*Hint* : use the `==` operator to compare the two vectors). How many are the same, that is, how many of the first letters are actually in alphabetical order already? -->
<!-- ```{r } -->
<!-- # Extract first character of each line -->
<!-- lyric1 <- substr(lyric,1,1) -->

<!-- # Sort and compare -->
<!-- sort(lyric1) == lyric1 -->

<!-- # And find if they are all the same -->
<!-- all(sort(lyric1) == lyric1) -->

<!-- # Count how many lines were already alphabetical -->
<!-- sum(sort(lyric1) == lyric1) -->
<!-- ``` -->

<!-- **1.** Find the most frequent word used in the lyrics. To do this, first paste all lyrics together into one string, then split the string into words, remove commas, then count the words. You will need to use a new function that we will see again in Section \ref{ \@ref(lapply}. *Hint* : use a combination of `paste`, `strsplit`, `gsub`, `table` and `sort`. -->

<!-- ```{r } -->
<!-- # Paste together all lyrics -->
<!-- lyr_all <- paste(lyric, collapse=" ") -->

<!-- # Remove commas -->
<!-- lyr_all <- gsub(",","",lyr_all) -->

<!-- # Split by space -->
<!-- lyr_words <- strsplit(lyr_all, " ") -->

<!-- # Sort frequencies in descending order and show only first few words -->
<!-- head(sort(table(lyr_words), decreasing=TRUE)) -->
<!-- ``` -->


### DNA Aerobics

DNA sequences can also be represented using text strings. In this exercise, you will make an artificial DNA sequence. 


**1.** Make a random DNA sequence, consisting of a 100 random selections of the letters `C`,`A`,`G`,`T`, and paste the result together into one character string (*Hint* : use `sample` as in Section  \@ref(randomnumbers) with replacement, and use `paste` as shown in Section  \@ref(workingtext). Write it in one line of R code.

```{r }
paste(sample(c("C","A","G","T"), 100, replace=TRUE), collapse="")
```




